---
title: "Least-Squares Analysis"
author: "your name"
format: html
editor: visual
execute:
  cache: false
  echo: true
---

## Getting started

Load the necessary R libraries

```{r results="hide"}
library(knitr)
library(tidyverse)
```

## Ordinary Least-Squares

### Example with fake data

Generate some data with known parameters:

```{r}
set.seed(12345)
b0 <- 3.5
b1 <- 7.2
sigma = 10
df <- tibble(x = seq(0, 10, length.out = 100),
             y = b0 + b1 * x + 
               rnorm(length(x), 0, sigma))
```

Use the `lm()` function to analyze the data using ordinary least-squares:

```{r}
fit <- lm(y ~ x, df)
```

Show a summary of the results:

```{r}
summary(fit)
```

We can extract coefficients from the fit using the `coef()` function:

```{r}
coef(fit)

coef(fit)["(Intercept)"]
coef(fit)["x"]
```

We can also get detailed information about the fit from the `summary()`
function:

```{r}
s <- summary(fit)

s$coefficients

s$coefficients['x', 'Estimate']
s$coefficients['x', 'Std. Error']
```

#### Plotting with Ordinary Least-Squares:

It's always useful to plot your data and the fit. We can use the
`geom_smooth()` function to add ordinary least-squares analysis to a plot of
the data:

```{r}
ggplot(df, aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = "lm")

```

The gray area around the fitted line shows the 95% confisence interval for the
line. This accounts for uncertainties in the intercept and slope of the line.

Remember that from the definition of confidence interval, this means that if
you performed this analysis with many different samples of data from the same
distribution, and plotted the confidence interval for each of these analysis,
we expect that 95% of these confidence intervals would contain the correct
line.

The correct line is determined by the distribution from which we sample the
data, so all of the samples have the same "correct line," but each will produce
a different estimate of the line and its confidence interval, because each
sample only looks at a finite amount of data from the true distribution.

### Ordinary Least-Squares with Real Data

Now, let's load some real data. Here, I'm taking paleoclimate data that uses
tree rings to predict summer air temperatures in the Mackenzie Delta in
Northwestern Canada.

The data comes from T.J. Porter *et al*., *Quaternary Res.* **80**, 167 (2013),
DOI:
[10.1016/j.yqres.2013.05.004](https://doi.org/10.1016/j.yqres.2013.05.004).

```{r}
paleo <- read_rds("tree-rings.rds")
glimpse(paleo)
```

Fit a straight line to the data using `lm()`

```{r}
fit_paleo <- lm(t_min ~ growth_index, paleo)
```

Summarize the ordinary-least-squares fit:

```{r}
summary(fit_paleo)
```

Plot the data

```{r}
ggplot(paleo, aes(x = growth_index, y = t_min)) + 
  geom_point() +
  geom_smooth(method="lm") +
  labs(x = "Tree ring growth index",
       y = expression(paste("Minimum summer temperature (", 
                            degree * C, ")")))
```

## General Least-Squares

But what do we do if the data doesn't fall on a straight line?

Load some different data. This is from measurements of ^40^Ar diffusion in
biotite. The researchers measured diffusion of ^40^Ar in different sized
biotite grains, at different temperatures and pressures.

The data come from T.M. Harrison *et al*., *Geomchim. Geocosm. Act.* **49**,
2461 (1985). DOI:
[10.1016/0016-7037(85)90246-7](https://doi.org/10.1016/0016-7037(85)90246-7)

```{r}
diffus <- read_rds("argon-diffusion.rds")
glimpse(diffus)
```

Plot the data:

```{r}
ggplot(diffus, aes(x = t, y = d)) +
  geom_point(size = 3) + 
  labs(x = "T(K)", y = "D (cm^2 / sec)")
```

Data points represent different grain sizees and pressures, so it would be
helpful to use other plot attributes to show these. In the language of the
Tidyverse and the `ggplot2` package, we call these different "aesthetics" of
the plot.

The grain sizes and pressures have distinct values. There are 4 grain sizes
(`r unique(diffus$a) |> sort() |> combine_words()` μm) and 2 pressures
(`r unique(diffus$P) |> sort() |> combine_words()` kbar), but R doesn't know
what we intend, so we convert the radius (`diffus$a`) and pressure (`diffus$P`)
to factors, which tells R that these variables have discrete values, and should
be plotted with discrete shapes and colors:

```{r}
diffus <- diffus |> mutate(P = factor(P), a = factor(a))

ggplot(diffus, aes(x = t, y = d)) + 
  geom_point(aes(color = P, shape = a), size = 3) + 
  scale_color_brewer(palette = "Dark2", name = "P (kbar)") + 
  scale_shape_manual(values = c("56" = 15, "80" = 16, 
                                "97" = 17, "202" = 18), 
                     name = expression(paste("radius (", 
                                             mu * m, ")"))) +
  labs(x = "T (K)", 
       y = expression(paste("D (", cm^2 / sec, ")")))

```

The data clearly don't lie along a straight line. Indeed, the physics of
diffusion tells us that we should expect a linear relationship between
$\log(D)$ and $1/T$.

Because the actual values of $T$ are close to 1000 Kelvin, we transform the
temperazture variable to $1000 / T$ so the values will be close to 1.

```{r}
diffus <- diffus |> mutate(t_inv = 1000 / t, 
                           log_d = log(d))
```

Now, plot the data. I'm going to use this plot again, later, so I save it to a
variable `p_xform`

```{r}
p_xform <- ggplot(diffus, aes(x = t_inv, y = log_d)) + 
  geom_point(aes(color = P, shape = a), size = 3) + 
  scale_color_brewer(palette = "Dark2", 
                     name = "P (kbar)") + 
  scale_shape_manual(values = c("56" = 15, "80" = 16, 
                                "97" = 17, "202" = 18), 
                     name = expression(paste("radius (", 
                                             mu * m, ")"))
                     ) +
  labs(x = expression(paste(1 / T, " (", 10^3 / K, ")")), 
       y = expression(paste(log~D, " (", cm^2 / sec, ")")),
       title = "Transformed Data") 

p_xform
```

This looks a lot more linear, except for the high-pressure (14 kbar) data.

Use `lm()` to fit the transformed data. In this way, our transformations
convert a general least-squares problem into an ordinary least-squares problem.
However, not all general least-squares problems can be transformed into
ordinary least-squares, so this is a special case.

$$
\log{D} = \beta_0 + \beta_1 \frac{1000}{T} + \varepsilon
$$

```{r}
fit_diffus <- lm(log_d ~ t_inv, diffus)
```

Now examine the summary

```{r}
summary(fit_diffus)
```

And repeat the plot, but add a least-squares analysis. Because I saved the plot
to a variable, all I have to do is add one new layer to the plot.

```{r}
p_xform + geom_smooth(method="lm")
```
